Title

### 2020-06-22 IWST/paper.tex
cut the rest of the paper, now only 9 pages

The first implementation, unigram-based sorting, is based on the 1-gram analysis, which means that we only take into consideration the actual token being completed. The implementation of the unigram model itself comes down to calculating individual token frequencies, i.e. the number of occurrences of each token in the source code. Then the completion candidates are sorted according to each one's frequency. In Listings you can see the frequencies data being passed to the sorter, and then the completions being sorted based on that information:

The second, more advanced implementation, is the bigram sorting strategy. As an n-gram model where n=2, it relies on both the token currently being completed and the history -- in this case, one token before the current one. To get the history word, I take the source code that is currently being edited (which is known to the context) and find the token preceding the one we are currently completing. After that, once the history word and each completion candidate are assembled into respective pairs of tokens, I calculate the probability of each of the n-gram sequences. Then the joint sequence probabilities are used for sorting each completion candidate (second word of the sequence). Below you can see how the sorting method looks in the end:

@@

The bigram model trained on source code data was implemented with the help of the NgramModel library\footnote{\url{https://github.com/pharo-ai/NgramModel}} available in Pharo. To calculate the probabilities, I created a bigram model, trained it on the source code tokens, cut off all the sequences with counts less than 50, and returned the result. Afterwards, the model was written to file (and then retrieved from file when I used it for sorting the results) -- this was done for speed purposes, described in detail in section.

When I first started using the NgramModel library, the functionality to reduce the number of n-gram sequences based on frequencies, or the functionality to read and write the model to file was not present. In the process of using this library for creating the bigram sorter, I extended it with the methods needed, as it might be useful for others to have this in place, too. Thus, through 4 pull requests I:
\begin{itemize}
    \item added the functionality to filter the table of n-grams (i.e. to cut off n-gram sequences with counts below a certain threshold)
    \item added writing to and reading n-gram models from file
    \item improved test coverage by writing tests for some additional examples
\end{itemize}

@@

The approach to improve the time efficiency was to set a certain threshold for the number of occurrences and to cut off all the tokens with frequencies below it. I put the threshold equal to 10, meaning if the token occurred less than 10 times throughout the whole history, it was irrelevant enough to be discarded. This way, I was able to cut off most of the miscellaneous, rarely encountered tokens, and shortened the dataset from 150k to just 16k entries. This made the frequencies lookup during completion sorting very fast and, as a result, it was now possible to type and use completion information without any delays (see the improvement in Figure).

@@

\subsection{Using the N-gram Sorters}
It is worth noting that both the n-gram sorters can actually be used in the Pharo IDE by loading the CompletionSorting library\footnote{\url{https://github.com/myroslavarm/CompletionSorting}}, which contains the implementation of this experiment. The unigram-based sorter is available as the FrequencyCompletionSorter, and the bigram-based as the BigramCompletionSorter -- they can be used by enabling the respective sorting option in the Settings.

The unigram sorter is quite fast and accurate. However, the bigram sorter is too slow for comfortable usage, and re-engineering the implementation to be faster is left for future work.

\section{Evaluation}
\subsection{Evaluation Overview}
Evaluation is an essential part of any research process. In the Richard Feynman quote above, the evaluation \textit{is} the experiment he is referring to. To paraphrase and make it more relevant to software engineering, it does not matter what kind of tool you implemented and how well do you think it works, there need to be clear metrics the implementation is measured on and a clear definition of what is considered to be a successful evaluation.

When it comes to scientific research, there are two main types of evaluation one can perform: quantitative and qualitative.

Quantitative, as can be guessed from its name, refers to the type of an evaluation experiment that is based on quantifiable data and uses objective metrics. This is the predominant evaluation approach in the natural sciences and computer science, i.e. measuring physical parameters in the case of the former, or testing performance scores and accuracy in the latter.

Qualitative methods, more native to fields such as sociology or psychology, are now being increasingly used in the evaluation of computer systems and software tooling\footnote{\cite{Hazz06a} in their paper discuss the qualitative approach and its relationship to the quantitative one.}. The main objective is to get the "feel" for the system and deeper explore the human interaction experience, with the data gathered primarily from observations and interviews and subsequently analysed.

Conventionally, a proper qualitative evaluation\footnote{\cite{Kapl05a} describe an extensive qualitative evaluation example and talk about the key benefits and considerations of this approach.} requires an extensive set-up, involving multiple people, recording their coding process, conducting feedback interviews, and so on. However, due to lack of time, the qualitative evaluation I conducted for this project is more minimal. Mainly, it involves manually selecting a number of common use cases and interacting with them as a user normally would, while recording the results and their meaning.

@@

Finally, this is all a matter of correctly managing the pros and cons of the evaluation. There are multiple ways to perform it, and multiple reasons to do something one way or another. The knowledge, interests, perceptions and skills of the researches undoubtedly play a role in the evaluation, and the subjective nature of the process is an undeniable part of the experiment and needs to be carefully addressed. In the end, the goal is to represent the process as closely as possible to how it really happens and report the results that are suitably meaningful, while exercising a healthy amount of unbiased professional detachment.

For this project both the quantitative and the qualitative approaches were used, as the first provides concise numeric results of the completion sorting performance, and the second allows to analyse common cases and demonstrates the actual human-computer interaction as it would happen with regular users (developers using the Pharo IDE). The next two sections detail the experiments and their results.

\subsection{Quantitative Evaluation}
The quantitative evaluation approach performed for this project is based on the paper by \cite{Robb08a}, wherein they used the change history data to evaluate the performance of their completion.

The idea with this approach is to recreate the coding process as it would happen naturally, i.e. a gradual appearance of new code as if it is being entered at the moment, and test the completion during that. Just as if a developer is typing and invoking a code completion engine at every step, with this approach the completion gets triggered for every token with the cursor between the 2nd and the 8th position (the range is taken from the \cite{Robb08a} paper). We can then compare whether the result suggested at that step matches the one that followed in real history.

@@

\begin{table}[H]
    \centering
    \begin{tabular}{lrrrr}
    \hline
    \textbf{Class} & \textbf{\# of Methods} & \textbf{Alphabetic} & \textbf{Unigram} & \textbf{Bigram} \\ \hline
    OrderedCollection & 64 & 0.32 & 0.38 & 0.30 \\ 
    CompletionEvaluation & 9 & 0.24 & 0.25 & 0.22 \\ 
    FrequencyCompletionSorter & 2 & 0.48 & 0.58 & 0.47 \\ 
    NgramModel & 21 & 0.32 & 0.36 & 0.29 \\ \hline
    \end{tabular}
\caption{Quantitative evaluation: the average success ratio (accuracy) for each class and sorting strategy}
\label{table:quan1}
\end{table}

From the results, we can see that the unigram sorter performs better than the alphabetic sorter, as expected. However, surprisingly, the bigram sorter performs both worse than the unigram sorter and even slightly worse than the alphabetic one.

@@

@@

It is worth noting that all the results are only recorded after two symbols were typed in, and having correct (most relevant) suggestions with such a short prefix appear among the top 10 results can be considered best-case scenario (as the correct suggestions would be more likely to be positioned higher with a longer prefix). Ergo, when the desired suggestion is missing from the top 10, it does not mean a complete failure of the completion engine or a sorting strategy used but rather hints that there is more work required to get to the correct suggestions, such as scrolling or typing more characters. However, as making the results as effective as possible is the goal of improving code completion in the first place, consistent placement of relevant results in the top 10 even after two symbols is a considerable advantage and does demonstrate which sorting strategy performs better.

Below you can see the 10 examples tested, the results, and their explanations.

\begin{description}
\item [Example 1:] \hfill
\begin{lstlisting}
    col := OrderedCollection new.
    col ad
\end{lstlisting}
This was typed into the Playground and the results were recorded with the cursor directly after \textit{ad}. For this context, the completion suggestions a developer would expect would be \textit{add:} and \textit{addAll:}. Here are the actual results:
\begin{table}[H]
    \centering
    \begin{tabular}{llr}
    \hline
    \textbf{Sorter Type} & \textbf{Completion Suggestion} & \textbf{Position} \\ \hline
    Alphabetic & - & - \\ \hline
    Unigram & \begin{tabular}[c]{@{}l@{}}add:\\ addAll:\end{tabular} & \begin{tabular}[c]{@{}r@{}}1\\ 2\end{tabular} \\ \hline
    Bigram & - & - \\ \hline
    \end{tabular}
\caption{Qualitative evaluation: the results of completing a message send in the Playground}
\label{table:qual1}
\end{table}
As we can see in the Table~\ref{table:qual1}, both the alphabetic and the bigram sorters did not manage to provide any relevant results in the top 10 suggestions, whereas the result from the unigram turned out exactly as expected.

@@

To briefly sum up the results of the manual qualitative evaluation, the unigram sorter still seems to be the best one out of the three, the same as in the quantitative evaluation. However, as seen in these examples, the bigram sorter is not very far behind, also more often than not prioritising the desired completions, which disagrees with the quantitative evaluation. So, while based on raw data the bigram sorter is the worst of the three \textit{on average}, from the manual evaluation it seems that it gives perfectly adequate results for examples that would be commonly encountered in Pharo. Thus, as part of the future work, it would be useful to investigate what exactly are the cases decreasing the bigram sorter's average accuracy, or, in other words, to find out what is the issue and what can be done to fix it.

@@

\subsection{Discoveries}
Through completing this work, I am now able to answer the research questions initially stated in Section \ref{sec:Introduction-Approach}:
\begin{RQ}
    \item \textbf{Can we improve the accuracy of code completion in the Pharo IDE by sorting candidate completions with n-gram language models?} As a result of the evaluation performed, the unigram based sorter has been shown to have a significantly better result than the default sorter in the Pharo IDE. The implementation is also fast enough for comfortable developer usage and is available by loading the CompletionSorting library available at \url{https://github.com/myroslavarm/CompletionSorting}.
    \item \textbf{How can we effectively evaluate the results of code completion enhanced by different sorting strategies?} For the quantitative evaluation, inspired by \cite{Robb08a}, I simulated the completion process as it would happen naturally, by generating source code sequences at various stages of typing, and comparing the results proposed to the ones in the codebase.

@@

\subsection{Directions of Future Work}
As can be seen in eval, the bigram sorter did not perform as well as expected. Contrary to the intuition that the higher order of n-gram should work better, the bigram performed much worse than the unigram. Additionally, it also seemed to give less relevant suggestions than the alphabetic sorter.

Hence, for future work, it would be useful to see what exactly went wrong with the implementation. It could be that there is a mismatch between how source code is parsed and tokenised in the training and test data. Or it could be a matter of managing the delimiters incorrectly, as punctuation in source code has a significant influence on the contextual meaning of any part of code. It could also be an issue with the NgramModel library, which was only used for training the bigram model, whereas the unigram model was implemented using a different approach.

In any case, this would be a good idea for the continuation of this project: analyse the issues of the existing bigram implementation and try to fix them and improve the performance of the bigram-based sorter.

For the quantitative evaluation, it would be useful to evaluate the results more precisely. Meaning that instead of an average accuracy, perhaps a more sophisticated formula could be used. For instance, the idea by \cite{Robb08a} involves making a note of the exact positions and prefix lengths and ranking completions, which give more relevant results for shorter prefixes, higher.

For the qualitative approach, a more extensive case study with multiple participants testing the actual completion in the IDE and reporting their feedback would be a good next step towards a more thorough evaluation of the results.

>>>>

The first implementation, unigram-based sorting, is based on the 1-gram analysis, which means that we only take into consideration the actual token being completed. The implementation of the unigram model itself comes down to calculating individual token frequencies, i.e. the number of occurrences of each token in the source code. Then the completion candidates are sorted according to each one's frequency.

The second, more advanced implementation, is the bigram sorting strategy. As an n-gram model where n=2, it relies on both the token currently being completed and the history -- in this case, one token before the current one. To get the history word, I take the source code that is currently being edited (which is known to the context) and find the token preceding the one we are currently completing. After that, once the history word and each completion candidate are assembled into respective pairs of tokens, I calculate the probability of each of the n-gram sequences. Then the joint sequence probabilities are used for sorting each completion candidate (second word of the sequence).

@@

The bigram model trained on source code data was implemented with the help of the NgramModel library\footnote{\url{https://github.com/pharo-ai/NgramModel}} available in Pharo. To calculate the probabilities, I created a bigram model, trained it on the source code tokens, cut off all the sequences with counts less than 50, and returned the result. Afterwards, the model was written to file (and then retrieved from file when I used it for sorting the results) -- this was done for speed purposes.

@@

The approach to improve the time efficiency was to set a certain threshold for the number of occurrences and to cut off all the tokens with frequencies below it. I put the threshold equal to 10, meaning if the token occurred less than 10 times throughout the whole history, it was irrelevant enough to be discarded. This way, I was able to cut off most of the miscellaneous, rarely encountered tokens, and shortened the dataset from 150k to just 16k entries. This made the frequencies lookup during completion sorting very fast and, as a result, it was now possible to type and use completion information without any delays.

@@

\subsection{Conclusion}
Both the n-gram sorters can actually be used in the Pharo IDE by loading the CompletionSorting library\footnote{\url{https://github.com/myroslavarm/CompletionSorting}}, which contains the implementation of this experiment. The unigram-based sorter is available as the FrequencyCompletionSorter, and the bigram-based as the BigramCompletionSorter -- they can be used by enabling the respective option in the Settings.

The unigram sorter is quite fast and accurate. The bigram sorter is still too slow for comfortable usage, and improving the implementation is left for future work.

\section{Evaluation}
\subsection{Evaluation Overview}
Evaluation is an essential part of any research process. When it comes to scientific research, there are two main types of evaluation one can perform: quantitative and qualitative.

Quantitative, as can be guessed from its name, refers to the type of an evaluation experiment that is based on quantifiable data and uses objective metrics. This is the predominant evaluation approach in the natural sciences and computer science, i.e. measuring physical parameters in the case of the former, or testing performance scores and accuracy in the latter. Qualitative methods, more native to fields such as sociology or psychology, are now being increasingly used in the evaluation of computer systems and software tooling \cite{Hazz06a}. The main objective is to get the "feel" for the system and deeper explore the human interaction experience, with the data gathered primarily from observations and interviews and subsequently analysed.

Conventionally, a proper qualitative evaluation \cite{Kapl05a} requires an extensive set-up, involving multiple people, recording their coding process, conducting feedback interviews, and so on. However, due to lack of time, the qualitative evaluation I conducted for this project is more minimal. Mainly, it involves manually selecting a number of common use cases and interacting with them as a user normally would, while recording the results and their meaning.

@@

Finally, this is all a matter of correctly managing the pros and cons of the evaluation. There are multiple ways to perform it, and multiple reasons to do something one way or another. For this project both the quantitative and the qualitative approaches were used, as the first provides concise numeric results of the completion sorting performance, and the second allows to analyse common cases and demonstrates the actual human-computer interaction as it would happen with regular users (developers using the Pharo IDE). The next two sections detail the experiments and their results.

\subsection{Quantitative Evaluation}
The quantitative evaluation approach performed for this project is based on the paper by Robbes et al. \cite{Robb08a}, wherein they used the change history data to evaluate the performance of their completion.

The idea with this approach is to recreate the coding process as it would happen naturally, i.e. a gradual appearance of new code as if it is being entered at the moment, and test the completion during that. Just as if a developer is typing and invoking a code completion engine at every step, with this approach the completion gets triggered for every token with the cursor between the 2nd and the 8th position (the range is also from \cite{Robb08a}). We can then compare whether the result suggested at that step matches the one that followed in real history.

@@

\begin{table}[H]
    \centering
    \small
    \setlength\tabcolsep{2pt}
    \begin{tabular}{lrrr}
    \hline
    \textbf{Class} & \textbf{Alphabetic} & \textbf{Unigram} & \textbf{Bigram} \\ \hline
    OrderedCollection & 0.32 & 0.38 & 0.30 \\ 
    CompletionEvaluation & 0.24 & 0.25 & 0.22 \\ 
    FrequencyCompletionSorter & 0.48 & 0.58 & 0.47 \\ 
    NgramModel & 0.32 & 0.36 & 0.29 \\ \hline
    \end{tabular}
\caption{Quantitative evaluation: the average success ratio (accuracy) for each class and sorting strategy}
\label{table:quan1}
\end{table}

From the results, we can see that the unigram sorter performs better than the alphabetic sorter, as expected. However, the bigram sorter performs both worse than the unigram sorter and even slightly worse than the alphabetic one.

@@

All the results are only recorded after two symbols were typed in, and having correct (most relevant) suggestions with such a short prefix appear among the top 10 results can be considered best-case scenario (as the correct suggestions would be more likely to be positioned higher with a longer prefix). Ergo, when the desired suggestion is missing from the top 10, it does not mean a complete failure of the completion engine or a sorting strategy used but rather hints that there is more work required to get to the correct suggestions, such as scrolling or typing more characters.

@@

\begin{description}
\item [Example 1:] \hfill
\begin{lstlisting}
    col := OrderedCollection new.
    col ad
\end{lstlisting}
This was typed into the Playground and the results were recorded with the cursor directly after \textit{ad}. For this context, the completion suggestions a developer would expect would be \textit{add:} and \textit{addAll:}. Here are the actual results:
\begin{table}[H]
    \centering
    \small
    \setlength\tabcolsep{2pt}
    \begin{tabular}{llr}
    \hline
    \textbf{Sorter Type} & \textbf{Completion Suggestion} & \textbf{Position} \\ \hline
    Alphabetic & - & - \\ \hline
    Unigram & \begin{tabular}[c]{@{}l@{}}add:\\ addAll:\end{tabular} & \begin{tabular}[c]{@{}r@{}}1\\ 2\end{tabular} \\ \hline
    Bigram & - & - \\ \hline
    \end{tabular}
    \caption{Qualitative evaluation: the results of completing a message send in the Playground}
\label{table:qual1}
\end{table}
As we can see in the Table~\ref{table:qual1}, both the alphabetic and the bigram sorters did not manage to provide any relevant results in the top 10 suggestions, whereas the result from the unigram turned out exactly as expected.

@@

To briefly sum up the results of the manual qualitative evaluation, the unigram sorter still seems to be the best one out of the three, the same as in the quantitative evaluation. However, as seen in these examples, the bigram sorter is not very far behind. Thus, as part of the future work, it would be useful to investigate what exactly are the cases decreasing the bigram sorter's average accuracy, or, in other words, to find out what is the issue and what can be done to fix it.

@@

\subsection{Discoveries}
Through completing this work, I am now able to answer the research questions that were stated at the beginning:
\begin{RQ}
    \item \textbf{Can we improve the accuracy of code completion in the Pharo IDE by sorting candidate completions with n-gram language models?} As a result of the evaluation performed, the unigram based sorter has been shown to have a significantly better result than the default sorter in the Pharo IDE. The implementation is also fast enough for comfortable developer usage and is available by loading the CompletionSorting library available at \url{https://github.com/myroslavarm/CompletionSorting}.
    \item \textbf{How can we effectively evaluate the results of code completion enhanced by different sorting strategies?} For the quantitative evaluation, inspired by \cite{Robb08a}, I simulated the completion process as it would happen naturally, by generating source code sequences at various stages of typing, and comparing the results proposed to the ones in the codebase.

@@

\subsection{Directions of Future Work}
As can be seen in the previous section, the bigram sorter did not perform as well as expected. Contrary to the intuition that the higher order of n-gram should work better, the bigram performed much worse than the unigram. Additionally, it also seemed to give less relevant suggestions than the alphabetic sorter. Thus, it would be useful to analyse the issues of the existing bigram implementation and try to fix them and improve the performance of the bigram-based sorter.

For the quantitative evaluation, I want to evaluate the results more precisely: instead of an average accuracy, perhaps a more sophisticated formula could be used. For instance, the idea by Robbes et al. \cite{Robb08a} involves making a note of the exact positions and prefix lengths and ranking completions, which give more relevant results for shorter prefixes, higher. For the qualitative approach, a more extensive case study with multiple participants testing the actual completion in the IDE and reporting their feedback would be a good next step towards a more thorough evaluation of the results.
